Well-quasi-orderings on word languages

### 2024-10-17 src/infixes.tex
restructuring paper

% LTeX: language=en-GB 
\section{Infixes and Regular Languages}
\label{infixes-regular:sec}

@@

\section{Bounded Languages and Amalgamation Systems}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument on regular langugages stated in
\cref{pumping-periods:lem}. Our first remark is that this idea can
be generalized to the larger class of \intro{bounded languages}, i.e.,
languages $L \subseteq \Sigma^*$ such that there exists words $w_1, \dots, w_n$
satisfying $L \subseteq w_1^* \cdots w_n^*$.

@@

    Furthermore, $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
    Indeed, consider a sequence $\seqof{u_i}$ of words in $L$. Without loss of
    generality, we may consider a subsequence such that $\card{u_i} <
    \card{u_{i+1}}$ for all $i \in \Nat$. Because $\ThueMorse$ is \kl{uniformly
    recurrent}, there exists $k \geq 1$ such that $u_1$ is an \kl{infix} of
    every word $v$ of size at least $k$. In particular, $u_1$ is an \kl{infix}
    of $u_k$, hence the sequence $\seqof{u_i}$ is \kl(wqo){good}.

@@

\subsection{Amalgamation Systems}
\label{infixes-amalgamation:subsec}

It turns out that there is a rather large family of systems for which pumping
arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Sch\"{u}tze, and Zetzsche \cite{ASZZ24}. Having done the heavy lifting on
\kl{bounded languages}, the rest of this section is mostly devoted to the
introduction of \kl{amalgamation systems} and collecting the necessary pumping
argument they enjoy. Using this meta-proof, we will generalize
\cref{infix-finite-automata:thm} to context-free grammars, languages recognized
by vector addition systems with state (VASS), and more. The goal of this
section is not to introduce all of these systems, or justify their usefulness,
but to state and prove the following theorem.

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    \proofref{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

Let us now formally introduce the notion of \kl{amalgamation systems}, and
recall some results from \cite{ASZZ24} that will be useful for the proof of
\cref{infix-amalgamation:thm}. The notion of \kl{amalgamation system} is
tailored to produce \emph{pumping arguments}, which is exactly what our
\cref{pumping-periods:lem} talks about. At the core of a pumping argument,
there is a notion of \emph{run}, which could for instance be a sequence of
transitions taken in a finite state automaton. Continuing on the analogy with
finite automata, there is a natural ordering between runs, i.e., a run is
smaller than another one if one can ``unroll loops" of the first run to obtain
the second one. Typical pumping arguments then rely on the fact that
\emph{minimal} runs are of finite size, and that all other runs are
obtained by ``gluing" minimal runs together. This is exactly what
\kl{amalgamation systems} are about.

\AP Let us recall that over an alphabet $(\Sigma, =)$ a \kl{subword embedding}
between two words $u \in \Sigma^*$ and $v \in \Sigma^*$ is a function $\rho
\colon \range{\card{u}} \to \range{\card{v}}$ such that $u_i = v_{\rho(i)}$ for
all $i \in \range{\card{u}}$. We write $\intro*\HigEmb(u,v)$ the set of all
\kl{subword embeddings} between $u$ and $v$. It may be useful to notice that
the set of finite words over $\Sigma$ forms a category when we consider
\kl{subword embeddings} as morphisms, which is a fancy way to state that
$\mathrm{id} \in \HigEmb(u,u)$ and that $f \circ g \in \HigEmb(u,w)$ whenever
$g \in \HigEmb(u,v)$ and $f \in \HigEmb(v,w)$, for any choice of words
$u,v,w \in \Sigma^*$.

\AP Given a \kl{subword embedding} $f \colon u \to v$ between two words $u$ and
$v$, there exists a unique decomposition $v = \GapWord{f}{0} u_1 \GapWord{f}{1}
\cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$ where $\GapWord{f}{i} =
v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$, $\GapWord{f}{k} =
v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$. We say that
$\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$. We encourage the
reader to look at \cref{gap-word-embedding:fig} to see an example of the
\kl{gap words} resulting from a \kl{subword embedding} between two words. These
\kl{gap words} will be useful to describe how and where runs of a system
(described by words) can be combined.

\begin{figure}
    \centering
    \includestandalone[width=\linewidth]{fig/gap-word-embedding-standalone}
    \caption{The \kl{gap words} resulting from a \kl{subword embedding} between two 
    finite words.}
    \label{gap-word-embedding:fig}
\end{figure}

\begin{figure}
    \centering
    \includestandalone[width=\linewidth]{fig/run-amalgamation-standalone}
    \caption{We illustrate how 
        embeddings $f$ and $g$ between runs of an
        \kl{amalgamation system} can be glued
        together, seen on their canonical decomposition.
    }
    \label{amalgamation-runs:fig}
\end{figure}

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings},
    and $\canrun \colon R \to (\Sigma \uplus \set{\varepsilon})^*$ is a 
    function computing a \emph{canonical decomposition} of a run.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{Word Embeddings as Morphisms.} For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item \emph{$(R, E)$ Forms a Category.}
            For all $r,s,t \in R$,
            $\mathrm{id} \in E(r,r)$,
            and whenever $f \in E(r,s)$ and $g \in E(s,t)$,
            then $g \circ f \in E(r,t)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \leq_E)$ is a well-quasi-ordered set,
            where $r \leq_E s$ if and only if $E(r,s) \neq \emptyset$.
        \item \emph{Controlled Amalgamation.}
            For all $r, s, t \in R$,
            for all $f \in E(r,s)$,
            for all $g \in E(r,t)$,
            and for all $0 \leq i_0 \leq \card{\canrun(r)}$,
            there exists a run $z \in R$ and morphisms
            $h \in E(s,z)$ and $k \in E(t,z)$
            satisfying
            $h \circ f = k \circ g$,
            $\GapWord{h \circ f}{i} = \GapWord{f}{i} \GapWord{g}{i}$
            or 
            $\GapWord{h \circ f}{i} = \GapWord{g}{i} \GapWord{f}{i}$
            for every $0 \leq j \leq n$,
            and
            $\GapWord{h \circ f}{i_0} = \GapWord{f}{i_0} \GapWord{g}{i_0}$.
            We refer to \cref{amalgamation-runs:fig} for an illustration 
            of this property.
    \end{enumerate}

    The \intro{amalgamation language} of such a system
    is the set of all words $w \in \Sigma^*$ such that
    there exists a run $r \in R$
    such that the concatenation of the letters of its
    canonical decomposition, written $\intro*\yieldrun(r)$,
    equals $w$.
\end{definition}

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing" of runs together to obtain new runs. Let us
recall some examples of languages that can be recognized by \kl{amalgamation
systems} given by the authors of the original paper: regular languages
\cite[Theorem 5.3]{ASZZ24}, VASS as a consequence of \cite[Theorem
5.5]{ASZZ24}, context-free languages as a consequence of \cite[Theorem
5.10]{ASZZ24}. For this paper to be self-contained, we will also recall how
runs of a finite state automaton can be understood as an \kl{amalgamation
system}.

\begin{example}[{\cite[Section 3.2]{ASZZ24}}]
    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
    \in Q \times \Sigma \times Q$,
    and $R \subseteq \Delta^*$ be the set of 
    words over transitions that start with the initial state $q_0$,
    end in a final state $q_f \in F$, and such that the end state of a
    letter is the start state of the following one.
    The canonical decomposition $\canrun \colon R \to \Sigma^*$
    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
    that maps $(q,a,p)$ to $a$.
    Finally, given two runs $r$ and $s$ of the automaton,
    we say that an embedding $f \in \HigEmb(\canrun(r), \canrun(s))$
    belongs to $E(r,s)$ when
    $f$ is also defining an embedding from $r$ to $s$ as words in $\Delta^*$,
    where $\Delta$ is equipped with the equality relation.

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    By definition, the embeddings inside $E(r,s)$ are defined as elements
    of $\HigEmb(\canrun(r), \canrun(s))$, and they compose properly.
    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
    according to Higman’s Lemma \cite{HIG52}.

    Let us now move to proving that the system satisfies the amalgamation
    property. Given three runs $r,s,t \in R$, and two embeddings $f \in E(r,s)$
    and $g \in E(r,t)$, we want to construct an amalgamated run $s \vee t$.
    Because letters in the run $r$ respect the transitions of the automaton
    (i.e., if the letter $i$ ends in state $q$, then the letter $i+1$ starts in
    state $q$), then the \kl{gap word} at position $i$ starts in state $q$ and
    ends in state $q$ too. This means that for both embeddings
    $f$ and $g$, the \kl{gap words} are read by the automaton by looping
    on a state. In particular, these loops can be taken in any order
    and continue to represent a valid run. That is, we can even select
    the order of concatenation in the amalgamation for \emph{all} 
    $0 \leq i \leq \card{\canrun(r)}$ and not just for one separately.

    Let us now remark that 
    the language of this amalgamation system is
    the set of $\yieldrun(r)$ when $r$ ranges over $R$,
    and because $R$ is the set of valid runs of the automaton,
    and $\yieldrun(r)$ is the word read along this run,
    we immediately conclude.
\end{proof}

Let us now introduce a combinatorial lemma that explains how \kl{gap languages}
can be pumped. Note that a single \kl{gap language} is always stable under
concatenation, and our concern will be on a potential description of
\emph{simultaneously} pumping different \kl{gap languages} to produce valid
runs.

\begin{lemma}
    \label{pumping-gap-languages:lem}
    Let $(\Sigma, R, E, \canrun)$ be an \kl{amalgamation system}, let $r \in R$
    be a run of size $n \in \Nat$, and let $s, t \in R$ be two runs
    with embeddings $f \in E(r,s)$ and $g \in E(r,t)$.
    Let $\seqof{u_i}[0 \leq i \leq n]$ be the sequence of \kl{gap words}
    for $f$, that is, $u_i =
    \GapWord{f}{i}$ for all $0 \leq i \leq n$. Similarly, let
    $\seqof{v_i}[0 \leq i \leq n]$ be the sequence of \kl{gap words} for $g$.

    Then, for all $k \in \Nat$, and $0 < \ell < n$, the following words belong to 
    the language of the \kl{amalgamation system}.
    \begin{enumerate}
        \item $(\prod_{i = 0}^{\ell - 1} u_i^{x_i} v_i u_i^{y_i} v_i u_i^{z_i}) 
               v_\ell u_\ell^k v_\ell
               (\prod_{i = \ell+1}^{n} u_i^{x_i} v_i u_i^{y_i} v_i u_i^{z_i})$,
        \item $(\prod_{i = 0}^{n - 1} u_i^{x_i} v_i u_i^{y_i}) 
               v_n u_n^k$,
        \item $u_0^k v_0 
            (\prod_{i = 1}^{n} u_i^{x_i} v_i u_i^{y_i})$.
    \end{enumerate}
    Where, for all $0 \leq i \leq n$, $x_i + y_i + z_i = k$, with $z_i = 0$
    in the two last items.
\end{lemma}
\begin{proof}
    The result immediately follows from the embedding property
    applied inductively on runs obtained by gluing 
    $k$ times $s$ with one or two times $t$ (depending on the item number).
\end{proof}

In the upcoming proof of \cref{infix-amalgamation:thm}, we will use the notion
of \intro{bounded language} for languages $L \subseteq \Sigma^*$ such that
there exists words $w_1, \dots, w_n$ satisfying $L \subseteq w_1^* \cdots
w_n^*$. This will be useful for us to derive the inclusion of a language $L$
into a union of \kl{chains}. Note that we isolate this intermediate result in a
separate theorem that could be applied beyond \kl{amalgamation systems}.

\begin{theorem}
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it is a finite union of \kl{chains}.
\end{theorem}
\begin{proof}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

    Now, admissible languages are ready to be pumped according to
    \cref{pumping-periods:lem}. For every admissible language,
    the size of a word that is not iterated is at most
    $N_0 \times m$ by definition, and we conclude that:
    \begin{equation}
    	\label{amalgamation-infix-wqo-inclusion:eqn}
        L \subseteq 
        \bigcup_{x,y \in \Sigma^{\leq n}}
        \bigcup_{u \in \Sigma^{\leq m \times N_0}}
        \InfPeriodChain{x} u \InfPeriodChain{y}
        \cup
        \InfPeriodChain{x} u
        \cup
        u \InfPeriodChain{x}
        \quad .
    \end{equation}
\end{proof}

To conclude this section and prove our main \cref{infix-amalgamation:thm}, it
suffices to demonstrate that languages recognized by \kl{amalgamation systems}
that are \kl{well-quasi-ordered} for the \kl{infix relation} are \kl{bounded
languages}.

\begin{lemma}
    \label{infix-amalgamation-bounded:lem}
    Let $L \subseteq \Sigma^*$ be a language recognized
    by an \kl{amalgamation system} that is \kl{well-quasi-ordered}
    for the \kl{infix relation}. Then $L$ is a \kl{bounded language}.
\end{lemma}
\begin{proof}
    Assume that $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
    and obtained by an \kl{amalgamation system}
    $(\Sigma, R, E, \canrun)$.

    Let us consider the set $M$ of minimal runs for the relation $\leq_E$,
    which is finite because the latter is a \kl{well-quasi-ordering}. By
    definition, for every word $w \in L$, there exists $r \in M$, $s \in R$,
    and $f \in E(r,s)$ such that $w = \canrun(s)$.
    In particular, we conclude that
    \begin{equation*}
        L \subseteq \bigcup_{r \in M} 
        \left(
        \GapLanguage{r}{0}
        \prod_{i = 1}^{\card{\canrun(r)}-1} \canrun(r)_i \GapLanguage{r}{i} 
        \right)
        \quad .
    \end{equation*}

    Let $u$ and $v$ be two words in the \kl{gap language}
    $\GapLanguage{r}{\ell}$ for some $r \in M$ and $0 < \ell <
    \card{\canrun(r)} \defined n$. One can apply \cref{pumping-gap-languages:lem}
    to conclude that  for all $k \geq 1$,
    \begin{equation*}
       \left(\prod_{i = 0}^{\ell - 1} u_i^{x_i} v_i u_i^{y_i} v_i u_i^{z_i}\right) 
       v u^k v
       \left(\prod_{i = \ell+1}^{n} u_i^{x_i} v_i u_i^{y_i} v_i u_i^{z_i}\right)
       \in 
       L
    \end{equation*}
    Where $x_i + y_i + z_i = k$ for all $0 \leq i \leq n$
    Let us assume without loss of generality that the sequence
    $\seqof{(x_i, y_i, n_i)}[i \geq 1]$ is pointwise increasing,
    and even increasing coordinate wise. This is possible because
    at least one of the coordinates tends to $+\infty$ and in the case
    of a variable that is bounded, one can extract the sequence to
    simply not use this variable.

    Now, leveraging \cref{pumping-periods:lem}, we get that for infinitely many
    $k \geq 1$, $u v^k u$ is an infix of some $x^p$ where $x$ has size smaller
    or equal than $u$ and $v$. This proves  that $u$ and $v$ are comparable for
    the prefix, infix, and suffix relations. In particular, we have can
    conclude that $\GapLanguage{r}{i}$ is a \kl{bounded language}, because of
    \cite[Lemma 4.1]{ASZZ24}, when $0 < i < \card{\canrun(r)}$. The proof goes
    on similarly for proving that $\GapLanguage{r}{0}$ and $\GapLanguage{r}{n}$
    are \kl{bounded languages}, using the other items of
    \cref{pumping-gap-languages:lem}.

    Because \kl{bounded languages} are stable under concatenation and finite
    unions, we conclude that $L$ itself must be a \kl{bounded language}.
\end{proof}

\begin{proofof}{infix-amalgamation:thm}
    We first apply \cref{infix-amalgamation-bounded:lem},
    and conclude with \cref{bounded-language:thm}.
\end{proofof}

For the next result, we impose a mild restriction by requiring the language class to be closed under rational transductions. This property holds for most common language classes, including all known \kl{amalgamation systems}.

\begin{theorem}
	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be a class of languages recognized by amalgamation systems and closed under rational transductions. Then the following are equivalent:

	\begin{enumerate}
		\item\label{wqo-infix-decidable} \kl[well-quasi-order]{Well-quasi-orderedness} of the \kl{infix relation} is decidable.
		\item\label{emptiness-decidable} Emptiness is decidable.
	\end{enumerate}
\end{theorem}

\begin{proof}
	\cref{emptiness-decidable} $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{amalgamation-infix-wqo-inclusion:eqn} effective.

	Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of the bounds $n$, $m$, and $N_0$, this language is regular. As $\mathcal{C}$ is closed under rational transductions, we can therefore reduce the inclusion to emptiness of $L \cap \overline{R}$. However, we need to find these bounds first.

	To determine values for $n$ and $m$, we first test if $L$ is bounded. As the algorithm in \cite[Section 4.2]{ASZZ24} is effective, this yields words $w_1, \ldots w_n$ such that $L \subseteq w_1^* \cdots w_n^*$ and therefore yields also the bounds in questions.

	To determine the value for $N_0$, we first computer maximal subsets $S \subseteq [n]$ such that $w_i$ for $i \in S$ are simultaneously unbounded. We do this by applying a transduction mapping $w_i$ to some fresh letter $a_i$ if $i \in S$ and $\varepsilon$ otherwise, and testing for simultaneous unboundedness \cite[Section 4.1]{ASZZ24}. Given a candidate bound $n_0$ for $N_0$, we construct the languages $L_{S,n_0} = w_1^{\circ_1}\cdots w_n^{\circ_n}$ where $\circ_i = *$ if $i \in S$ and $\circ_i = \leq n_0$ otherwise. We then test if $L \subseteq \bigcup_{S \text{ maximal}} L_{S,n_0}$. If yes, $n_0$ is our value for $N_0$, otherwise, we increase it and repeat the construction. As we chose maximal sets $S$, this procedure will eventually terminate.

	\cref{wqo-infix-decidable} $\Rightarrow$ \cref{emptiness-decidable}. We 
	consider the transduction $\mathcal T = \Sigma^* \times \{a, b\}$. Then 
	for any language $L \in \mathcal C$, $\mathcal TL$ is well-quasi-ordered 
	by infix if and only if $L$ is empty.
\end{proof}

Let us conclude this section by fully characterizing what are
actually the chains for the \kl{infix} relation, without any 
regularity assumption.

\begin{lemma}
    \label{chains-infix:lem}
    Let $L$ be a \kl{chain} for the \kl{infix relation}. Then, there exists
    a \kl{chain} $P$ for the \kl{prefix relation} and a \kl{chain}
    $S$ for the \kl{suffix relation} such that $L \subseteq SP$.
\end{lemma}
\begin{proof}
    Because $L$ is a chain,
    we have an enumeration $L = \seqof{w_n}$,
    where $s_n \defined \card{w_n}$ is increasing, 
    and functions
    $f_{i,j} \colon \set{1, \dots, s_i} \to \set{1, \dots s_{j}}$
    satisfying
    that $f_{j,k} \circ f_{i, j} = f_{i, k}$,
    $f_{i,i} = \mathrm{id}$, and
    $w_i = w_j[f_{i,j}(1):f_{i,j}(s_i)]$ for all $i \leq j$.
    Now, we can build the infinite word $w_\infty$ indexed by $\Rel$, and
    embeddings $f_{i, \infty} \colon \set{1, \dots, s_i} \to \Rel$ such that
    $w_i = w_\infty[f_{i, \infty}(1):f_{i, \infty}(s_i)]$, and $f_{i, \infty}
    \circ f_{j, i} = f_{j, \infty}$ for all $i \leq j$.

    Let us define $w_l$ as the (infinite) word $w_\infty[-\infty: f_{0,
    \infty}(s_0)]$ and $w_r$ as the (infinite) word $w_\infty[f_{0,
    \infty}(s_0): +\infty]$. It is now clear that every word $w_i$ can be
    written as $w_i = w_{i,l} w_{i,r}$, where $w_{i,l}$ is a prefix of $w_r$
    and $w_{i,r}$ is a suffix of $w_l$. Furthermore, the functions $f_{i,
    \infty}$ can be used to notice that $\seqof{w_{i,r}}$ is a \kl{chain} for
    the \kl{prefix relation}, and $\seqof{w_{i,l}}$ is a \kl{chain} for the
    \kl{suffix relation}, which allows us to conclude.
\end{proof}

>>>>

% LTeX: language=en-GB 
\section{Infixes and Bounded Languages}
\label{infixes-regular:sec}

@@

@@

    Furthermore, $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
    Indeed, consider a sequence $\seqof{u_i}$ of words in $L$. Without loss of
    generality, we may consider a subsequence such that $\card{u_i} <
    \card{u_{i+1}}$ for all $i \in \Nat$. Because $\ThueMorse$ is \kl{uniformly
    recurrent}, there exists $k \geq 1$ such that $u_1$ is an \kl{infix} of
    every word $v$ of size at least $k$. In particular, $u_1$ is an \kl{infix}
    of $u_k$, hence the sequence $\seqof{u_i}$ is \kl(sequence){good}.

@@

### 2024-10-17 src/prefixes.tex
restructuring paper

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the \kl{tree of prefixes} such that from every point of the branch, 
    one can reach a word in $L \setminus B$.
\end{definition}

Let us illustrate the notion of \kl{antichain branch} over the alphabet $\Sigma
= \set{a,b}$, and the language $L = a^* b$. In this case, the set $a^*$ (which
is a branch of the \kl{tree of prefixes}) is an \kl{antichain branch} for $L$.
This holds because for any $a^k$, the word $a \prefleq a^kb$ belongs to $L
\setminus a^*$. In general, the existence of an \kl{antichain branch} for a
language $L$ implies that $L$ contains an infinite \kl{antichain}, and because
the alphabet $\Sigma$ is assumed to be finite, one can leverage the fact that
the \kl{tree of prefixes} is finitely branching to prove that the converse
holds as well.

@@

One immediate application of \cref{antichain-branches-prefix:lem} is that
\kl{antichain branches} can be described inside the \kl{tree of prefixes} by a
monadic second order formula (\kl{$\MSO$-formula}), allowing us to leverage the
decidability of $\MSO$ over infinite binary trees \cite[Theorem 1.1]{RAB69}.

@@

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be an infinite language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists finite numbers $k,\ell \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $\ell$.
\end{corollary}

>>>>

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the \kl{tree of prefixes} such that from every point of the branch, 
    one can reach a word in $L \setminus B$. Formally:
    $\forall u \in B, \exists v \in \Sigma^*, uv \in L \setminus B$.
\end{definition}

Let us illustrate the notion of \kl{antichain branch} over the alphabet $\Sigma
= \set{a,b}$, and the language $L = a^* b$. In this case, the set $a^*$ (which
is a branch of the \kl{tree of prefixes}) is an \kl{antichain branch} for $L$.
This holds because for any $a^k$, the word $a^k \prefleq a^kb$ belongs to $L
\setminus a^*$. In general, the existence of an \kl{antichain branch} for a
language $L$ implies that $L$ contains an infinite \kl{antichain}, and because
the alphabet $\Sigma$ is assumed to be finite, one can leverage the fact that
the \kl{tree of prefixes} is finitely branching to prove that the converse
holds as well.

@@

One immediate application of \cref{antichain-branches-prefix:lem} is
that \kl{antichain branches} can be described inside the \kl{tree of prefixes}
by a monadic second order formula (\kl{$\MSO$-formula}), allowing us to
leverage the decidability of $\MSO$ over infinite binary trees \cite[Theorem
1.1]{RAB69}. This result will follow from our general decidability result in
\cref{amalgamation:sec} but serves as a good illustration.

@@

\begin{corollary}
    \label{prefixes-ordinal-invariants:cor}
    Let $L \subseteq \Sigma^*$ be a language that is
    \kl{well-quasi-ordered} by the \kl{prefix relation}. Then,
    \kl{maximal order type} of $L$ strictly smaller than $\omega^2$,
    the \kl{ordinal height} of $L$ is at most $\omega$, and
    its \kl{ordinal width} is finite.
\end{corollary}
